String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-06 2023-1-19/sections/overview.tex
overview

\zhilin{to be made consistent with section 6.}

@@

We first construct an automaton for the regular expression $(\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. We introduce three registers, say $r_1, r_2, r_3$, to represent the three counting operators.  
%Moreover, we introduce another register, say $r_0$, to denote the length of strings. 
Then we first construct a nondeterministic finite automaton (NFA) for $(\Sigma \setminus a)^* (\Sigma \setminus b)^* (\Sigma \setminus c)^*$, then add the updates of registers to transitions, resulting into a so-called cost-enriched finite automaton (CEFA, see Section~\ref{sec:automaton} for the definition) $\aut_1$,  as illustrated in Figure~\ref{fig:overview}(a). 
%Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
Furthermore, the counting bounds are specified by the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$. Note here we represent the counting operators symbolically by registers, instead of unfolding them explicitly. Moreover, we construct another CEFA $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)), which is actually an NFA, since $\Sigma^* c^+$ contains no (explicit) counting operators. Finally, we construct the third CEFA $\aut_3$ (see Figure~\ref{fig:overview}(c)), where a register $r_4$ is used to denote the length of strings and the accepting condition is $\ltrue$. 
%where the register $r_0$ is used to record the length of strings.

@@

\zhilin{stopped here}

Let $\tau_1, \cdots, \tau_m$ be an enumeration of the transitions of $\aut_1 \cap \aut_2 \cap \aut_3$ and $\myvec{v_1}, \cdots, \myvec{v_m}$ be the vectors of updates on the registers in these transitions. 
%For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
Then the satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$. 
From the results in \cite{SSMH04,VSS05}, an existential LIA formula $\psi(i_1, \cdots, i_m)$ can be computed to express the constraints on the number of occurrences of $\tau_1, \cdots, \tau_m$ in the accepting runs of $\aut_1 \cap \aut_2 \cap \aut_3$, where $i_1, \cdots, i_m$ are the integer variables representing the number of occurrences of $\tau_1, \cdots, \tau_m$ respectively. 
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-LIA}
\psi(i_1, \cdots, i_m) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [m]}  i_k v_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
\end{equation}
where for each $k \in [m]$, $v_k = (v_{k,1}, v_{k,2}, v_{k, 3}, v_{k, 4})$ is the vector of updates on the registers $r_0, r_1, r_2, r_3$ in the transition $\tau_k$, and $\alpha(r_0, r_1, r_2, r_3)$ is the accepting condition in $\aut_1 \times \aut_2$. Finally, the satisfiability of LIA formula can be solved by the off-the-shelf SMT solvers for LIA.

Nevertheless, since it is well known that the satisfiability of existential LIA formulas is NP-complete (see e.g. \cite{Haase18}), the performance could be improved if the size of the formula in~(\ref{eqn-LIA}) can be reduced. Since the size of the formula in~(\ref{eqn-LIA}) depends on the size of $\aut_1 \times \aut_2$, that is, the number of states, transitions, and registers in $\aut_1 \times \aut_2$, we propose the techniques to reduce the size of the CEFA $\aut_1 \times \aut_2$.

Our main idea to reduce the size of the CEFA $\aut_1 \times \aut_2$ is to \emph{ignore the characters in transitions}, since the updates of the values of registers in transitions are independent of the characters therein and the satisfiability of the formula in~(\ref{eqn-LIA}) only concerns about the values of registers. After the characters in transitions are ignored, each transition in $\aut_1 \times \aut_2$ is of the form $(q, \myvec{v}, q')$, where $\myvec{v}$ is the vector denoting the updates on the values of registers. Let $\cB$ denote the resulting automaton. We then think $\cB$ as an NFA, where the vectors $\myvec{v}$ are taken as the characters. Then we apply the determinization and minimization algorithm of NFAs to $\cB$, resulting into an automaton $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}. 
Note that $\cC$ contains only three states $q'_0, q'_1, q'_2$ and six transitions $\tau'_1, \cdots, \tau'_6$, thus containing much less number of transitions than $\aut_1 \times \aut_2$\footnote{This is the case even if we consider subsets of characters, e.g. $\{a\}, \{b\}, \{c\}, \Sigma \setminus \{a, b, c\}$, in the transitions of $\aut_1 \times \aut_2$.}.
Then we reinterpret $\cC$ as a CEFA with the acceptance condition $\alpha(r_0, r_1, r_2, r_3)$. 
Finally, we compute an LIA formula $\psi'(i_1, \cdots, i_6)$ from $\cC$ and reduce the satisfiability of the string constraint in~(\ref{eqn-running}) to the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-LIA-reduced}
\psi'(i_1, \cdots, i_6) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [6]}  i_k v'_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
\end{equation}
where $\myvec{v'_{k}}=(v'_{k,1}, v'_{k,2}, v'_{k,3}, v'_{k,4})$ is the vector of updates of registers of the transition $\tau'_k$ for $k \in [6]$.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.3\textwidth]{sections/overview-cefa-reduced.pdf}
  \caption{Reduced automaton $\cC$}
  \label{fig:overview:product:reduced}
\end{figure}
Furthermore, we can reduce the number of registers by merging several registers into one, if they are updated with the same value in every transition of $\aut_1 \times \aut_2$.

At last, for the string constraints that are satisfied by a collection of strings whose lengths are small, computing LIA formulas from CEFA and solving them by SMT solvers are still expensive.  
Therefore, we introduce a heuristic of searching for solutions based on \emph{under approximations}, before computing LIA formulas and solving them by SMT solvers. More precisely, we enforce a constant upper bound $B$ on the lengths of strings and search for a $B$-bounded satisfiable assignment, that is, a satisfiable assignment where the lengths of strings are bound by $B$. Initially we set $B$ to $1$. If there does not exist a $B$-bounded satisfiable assignment, then we let $B:=B+1$ and continue the search, until finding a satisfiable assignment or reaching a given time limit. For instance, if we remove $|x| > 120$ from the string constraint~(\ref{eqn-running}), then the resulting constraint is satisfiable. Setting the bound $B$ to $1$ is already sufficient to find a solution for $\aut_1 \times \aut_2$ (with $r_0 > 120$ ignored).

>>>>

%\zhilin{to be made consistent with section 6.}

@@

We first construct an automaton for the regular expression $(\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. We introduce three registers, say $r_1, r_2, r_3$, to represent the three counting operators.  
%Moreover, we introduce another register, say $r_0$, to denote the length of strings. 
Then we first construct a nondeterministic finite automaton (NFA) for $(\Sigma \setminus a)^* (\Sigma \setminus b)^* (\Sigma \setminus c)^*$, then add the updates of registers to transitions, resulting into a so-called cost-enriched finite automaton (CEFA, see Section~\ref{sec:automaton} for the definition) $\aut_1$,  as illustrated in Figure~\ref{fig:overview}(a). 
%Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
Furthermore, the counting bounds are specified by the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$. 
A string $w$ is accepted by $\aut_1$ if when reading the characters in $w$, $\aut_1$ applies the transitions to update the state and the values of registers, and reaches a final state $q$ in the end, moreover, the resulting values of the three registers, say $v_1, v_2, v_3$, satisfy the accepting condition. 
Note here we represent the counting operators symbolically by registers, instead of unfolding them explicitly. 
Moreover, we construct another CEFA $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)), which is actually an NFA, since $\Sigma^* c^+$ contains no (explicit) counting operators. Finally, we construct the third CEFA $\aut_3$ (see Figure~\ref{fig:overview}(c)), where a register $r_4$ is used to denote the length of strings and the accepting condition is $\ltrue$. (See Section~\ref{subsec:regex2cefa} for more details about the construction of CEFA.)
%where the register $r_0$ is used to record the length of strings.

@@

%For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
The satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$, that is, whether there exist $w \in \Sigma^*$ and $(v_1, v_2, v_3, v_4) \in \Int^4$ such that $w$ is accepted by $\aut_1 \cap \aut_2 \cap \aut_3$, producing a cost $(v_1, v_2, v_3, v_4)$ that satisfies $1 \le v_1 \le 60\wedge 1 \le v_2 \le 60 \wedge 0 \le v_3 \le 60$ and $v_4 > 120$. 
It is not hard to observe that the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is independent of the characters of $\aut_1 \cap \aut_2 \cap \aut_3$.  Therefore, the characters in $\aut_1 \cap \aut_2 \cap \aut_3$ can be ignored, resulting into an NFA $\cB$ over the alphabet $\costset$ (see Figure~\ref{fig:overview:product:reduced}(a)), where $\costset$ is the set of vectors from $\Int^4$ occurring in the transitions of $\aut_1 \cap \aut_2 \cap \aut_3$. Then the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is reduced to the problem of deciding whether there exists $w' \in \costset^*$ such that it is accepted by $\cB$ and its Parikh image (i.e. numbers of occurrences of characters), say $\eta_{w'}: \costset \rightarrow \Nat$, satisfies that $1 \le v'_1 \le 60\wedge 1 \le v'_2 \le 60 \wedge 0 \le v'_3 \le 60$ and $v'_4 > 120$, where $\myvec{v'} = \sum \limits_{\myvec{v} \in \costset} \eta_{w'}(\myvec{v}) \myvec{v}$. Intuitively, $\myvec{v'}$ here is a weighted sum of vectors $\myvec{v} \in \costset$, where the weight for $\myvec{v}$ is the number of occurrences of $\myvec{v}$ in $w'$. (See Section~\ref{subsec:cefadec} for more detailed arguments.)

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.8\textwidth]{sections/overview-cefa-reduced.pdf}
  \caption{Reduced automaton $\cB$ and $\cC$}
  \label{fig:overview:product:reduced}
\end{figure}

%Let $\tau_1, \cdots, \tau_m$ be an enumeration of the transitions of $\aut_1 \cap \aut_2 \cap \aut_3$ and $\myvec{v_1}, \cdots, \myvec{v_m}$ be the vectors of updates on the registers in these transitions.

%
Let $\costset = \{\myvec{v_1}, \cdots, \myvec{v_m}\}$. 
From the results in \cite{SSMH04,VSS05}, an existential LIA formula $\psi_\cB(\anivar_1, \cdots, \anivar_m)$ can be computed to define the Parikh image of strings that are accepted by $\cB$, where $\anivar_1, \cdots, \anivar_m$ are the integer variables to denote the number of occurrences of $\myvec{v_1}, \cdots, \myvec{v_m}$.
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-LIA}
\begin{array}{l}
\psi_\cB(\anivar_1, \cdots, \anivar_m) \wedge \bigwedge \limits_{1 \le j \le 4} r_j = \sum \limits_{1\le k \le m}  \anivar_k v_{k, j}\ \wedge \\
1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_4 > 120.
\end{array}
\end{equation}
%where for each $k \in [m]$, $v_k = (v_{k,1}, v_{k,2}, v_{k, 3}, v_{k, 4})$ is the vector of updates on the registers $r_0, r_1, r_2, r_3$ in the transition $\tau_k$, and $\alpha(r_0, r_1, r_2, r_3)$ is the accepting condition in $\aut_1 \times \aut_2$. 
Since the satisfiability of LIA formulas can be solved by the off-the-shelf SMT solvers for LIA, the satisfiability of the string constraint in~(\ref{eqn-running}) can thus be solved.

Nevertheless, it turns out that the algorithm illustrated above is insufficient to solve the string constraints with regex-counting and string-length efficiently, since the sizes of NFA obtained from CEFA can still be big, resulting into the LIA formulas of big sizes, while the satisfiability of existential LIA formulas is NP-complete (see e.g. \cite{Haase18}). 
To tackle this issue, we proposes techniques to reduce the sizes of NFA over the alphabet $\costset$.

%the performance could be improved if the size of the formula in~(\ref{eqn-LIA}) can be reduced. Since the size of the formula in~(\ref{eqn-LIA}) depends on the size of $\aut_1 \times \aut_2$, that is, the number of states, transitions, and registers in $\aut_1 \times \aut_2$, we propose the techniques to reduce the size of the CEFA $\aut_1 \times \aut_2$.

%Our main idea to reduce the size of the CEFA $\aut_1 \times \aut_2$ is to \emph{ignore the characters in transitions}, since the updates of the values of registers in transitions are independent of the characters therein and the satisfiability of the formula in~(\ref{eqn-LIA}) only concerns about the values of registers. After the characters in transitions are ignored, each transition in $\aut_1 \times \aut_2$ is of the form $(q, \myvec{v}, q')$, where $\myvec{v}$ is the vector denoting the updates on the values of registers. Let $\cB$ denote the resulting automaton. We then think $\cB$ as an NFA, where the vectors $\myvec{v}$ are taken as the characters.

We determinize $\cB$, apply the minimization algorithm to the resulting deterministic finite automaton (DFA), and obtain a DFA $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}(b). 
Note that $\cC$ contains only three states $q''_0, q''_1, q''_2$ and six transitions, while $\cB$ contains four states and eight transitions. 
Although the determinization and minimization may not reduce the sizes of NFA in the worst case, the experimental results in Section~\ref{sec:implementation} show us that they do reduce the sizes of NFA that are constructed from the regular expressions in practice. 
%thus containing much less number of transitions than $\aut_1 \times \aut_2$\footnote{This is the case even if we consider subsets of characters, e.g. $\{a\}, \{b\}, \{c\}, \Sigma \setminus \{a, b, c\}$, in the transitions of $\aut_1 \times \aut_2$.}.
%Then we reinterpret $\cC$ as a CEFA with the acceptance condition $\alpha(r_0, r_1, r_2, r_3)$. 
%Finally, we compute an LIA formula $\psi'(i_1, \cdots, i_6)$ from $\cC$ and reduce the satisfiability of the string constraint in~(\ref{eqn-running}) to the satisfiability of the following existential LIA formula
%\begin{equation}\label{eqn-LIA-reduced}
%\psi'(i_1, \cdots, i_6) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [6]}  i_k v'_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
%\end{equation}
%where $\myvec{v'_{k}}=(v'_{k,1}, v'_{k,2}, v'_{k,3}, v'_{k,4})$ is the vector of updates of registers of the transition $\tau'_k$ for $k \in [6]$.

%Furthermore, we can reduce the number of registers by merging several registers into one, if they are updated with the same value in every transition of $\aut_1 \times \aut_2$.

At last, for the string constraints that are satisfied by a collection of strings whose lengths are small, computing LIA formulas from NFA and solving them by SMT solvers are still expensive.  
Therefore, we introduce a heuristic of searching for solutions based on \emph{under approximations}, before computing LIA formulas and solving them by SMT solvers. More precisely, we enforce a constant upper bound $B$ on the lengths of strings and search for a $B$-bounded satisfiable assignment, that is, a satisfiable assignment where the lengths of strings are bound by $B$. Initially we set $\ell$, an integer variable to denoting the lengths of strings, to $1$. If there does not exist a $\ell$-bounded satisfiable assignment, then we let $\ell:=\ell+1$ and continue the search, until finding a satisfiable assignment or reaching the bound $B$. For instance, if we remove $|x| > 120$ from the string constraint~(\ref{eqn-running}), then the resulting constraint is satisfiable. Setting the bound $B$ to $3$ is sufficient to find a solution, which is witnessed by the following accepting run of $\cC$: $q^{\prime\prime}_0 \xrightarrow{(1,0,0,1)} q^{\prime\prime}_0 \xrightarrow{(0,1,0,1)} q^{\prime\prime}_1 \xrightarrow{(0,0,0,1)} q^{\prime\prime}_2$. Note that the accepting run produces $(1,1,0,3)$, a vector of register values,  which satisfies $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$.
